Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-28 introduction_new.tex

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current impact of pandemics is on average of over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status of the population at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contact, the probability of transmission upon contact, and the proportion of the population that is susceptible.
Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
Therefore, $\Re$ informs the strength of intervention required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections that would be generated by a single infection at time $t$ if in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: the number ofindividuals with immunity, higher in 2021; and the different measures already (?) in place when the lockdown was implemented???
$\Rc$ quantifies the latter effect.

These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a \emph{growth rate} $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on whether how an epidemic is changing.
The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence}, the number of new infections in a given time period  is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths) when combined with severity measures, such as the infection fatality rate (IFR), the proportion of infected individuals who die.

\emph{Transmission} refers to the process by which an infectious individual generates new infections. Estimating transmission refers to inferring this process. Understanding components of transmission  {\it{e.g.}} the number of contacts each individual makes per day) allow interventions to be targeted.
and to forecast future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time .
Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
Prevalence surveys, designed  to perform diagnostic testing of a sample of the population,  can be used to estimate the population prevalence,  but are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
Hence, they are typically estimated from other data sources through appropriate inferential methodologies.
{\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
Chapman & Hall/CRC, 2019.}}

@@

Data sources to inform outbreak surveillance are often observational in nature, {\it i.e.} they are not the output of designed studies.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

@@

Incidence-type data originates from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, as the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

Prevalence-type data derives from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
For example, the number of individuals in hospitalwith the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence surveys are one source of prevalence-type data that can be unbiased.
For instance, a representative proportion of the population (or a sub-population of interest) might be tested for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

On the other hand, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

@@

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Specific studies, {\it e.g.} longitudinal studies that follow  individuals over time, allow  estimation of estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately powered, which are expensive to run.

@@

The models with the weakest assumptions are the so-called \emph{phenomenological} models (add a reference!! Still Becker?).
These make almost no assumptions about the transmission process, viewing incidence as resulting from a point process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
A common model is the Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These include components of the transmission process  but do not explicitly describe the process.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic} of the most form:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions to incorporate heterogeneity, for example, to account for the age structure and the interaction between age groups.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Observational models} \label{intro:sec:disease-model}
 The infection process is hardly ever directly observed. However, as seen in previous sections,  indirect observations are available on the infection's sequelae, which can be linked to the latent infection process through appropriate observational models.
At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
This equation describes the expectation of an infectious sequelae process ({\it e.g.} doctor's consultation, death) $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability distribution of the time from infection to observation.
$f$ is an example of a \emph{delay function}, and  
$f$ is commonly the pdf of a distribution, or a transformation of it.

@@

Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
Typically, this is either the time to an event or the duration of a disease stage.
Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.  This generally comes from longitudinal studies where the occurrence of the infection's sequelea is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
Concerted Action (2000). Time from HIV- 1
seroconversion to AIDS and death in 13,000 individuals before the widespread use of
highly active anti-retroviral therapy: effect of age at seroconversion and exposure
category. Lancet 355, 1131-1137.)

@@

For large parts of the pandemic, community testing was widely available in the UK.
However,
the number of tests carried out varied over time and space, for a variety of reasons, including: policy changes on who was eligible for testing;availability of tests; and willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests did not necessarily reflect changes in disease incidence.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.

@@

However, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this, they are considered the gold-standard metric.

Seroprevalence surveys were conducted throughout the epidemic, the largest of which was testing donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and attempt to observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the  Coronavirus (COVID-19) Infection Survey (CIS), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

Many ad hoc studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
One of the study, which  I will use in what follows, is the Assessment of Transmission and Contagiousness of COVID-19 in Contacts (ATACCC) study. ~\autocite{singanayagamCommunity,hakkiOnset}. InATACCC  daily testing of a small number of individuals with short follow-up was performed.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability. However, the results from ATACCC are of limited value because of the small sample size and, particularly, of the short follow-up, which  prohibit estimation of the tails of these distributions.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of positivity.
However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

\subsection{Models}

@@

During the SARS-CoV-2 pandemic, analysis of the unprecedently available data was carried out in haste, as results were needed at pace. Therefore, a deep understanding of the potential of each data source was not possible.
In this thesis I investigate the potential of the CIS as a sinlge source of information for incidence and transmission estimation.

In the following subsection, I introduce the most relevant related work, carried out  both prior to the pandemic and during its course.

@@

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} uses a Poisson process as the infection model and a delay function, $f$, the distribution of the time between infection and symptomatic disease (incubation period); and an observation model where all diagnoses are observed.

In (see \cref{E-inc-prev}) I will adapt this backcalculation framework to data from the CIS, which is of a prevalence-type rather than incidence-type.
This difference induces a dependency structure between the observations on each day.
The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.

@@

In the application of the backcalculation to CIS data, the delay function $f$ is represented by the duration of positivity. Many attempts have been made to estimate its distribution. An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}, with 
the longest reported duration at 83 days, also reporting lack of virus detection beyond day 9, indicating that the individuals testing positive for longer than this were not infectious.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of positivity.
However, the study's cohort consisted only of healthcare workers, {it i.e.} working-age adults.

\Textcite{binnySensitivity} applied a slight variant of the methodology  in \textcite{hellewellPCRSensitivity} to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy in the earlier part of the pandemic, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.

>>>>

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it includes the effects of the behaviour and immunity status of the population at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
Therefore, $\Re$ informs the strength of intervention required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections that would be generated by a single infection at time $t$ if in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.

To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
The original and Alpha variants have different values for $\R$, isolating effect 1.
$\Rc$ differs during the two periods, including effects 1 and 2.
$\Re$ includes all these effects, as well as any others.

These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a \emph{growth rate} $r$.
For a given disease and population, a growth rate can be defined corresponding to each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.

\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
It is the primary metric for the magnitude of a pandemic .
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.

\emph{Transmission} refers to the process by which an infectious individual generates new infections.
Estimating transmission refers to inferring this process. Understanding this process, \eg the number of contacts each individual makes per day, allows targeting interventions and forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time (see \cref{E-inc-prev:sec:definitions} for a formal definition).
Whether a given individual is \emph{prevalent}, \ie has the disease, is often measurable through diagnostic testing.
Prevalence can be estimated through a \emph{prevalence survey}.
A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that \emph{directly} inform inference of the quantities in this section is challenging or prohibitively expensive.
Hence, they are typically estimated from other data sources through appropriate ???? methodologies.
{\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
Chapman & Hall/CRC, 2019.}}

@@

Data sources to inform outbreak surveillance are often observational in nature, \ie they are not the output of designed studies.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

@@

Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, such as the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
For example, the number of individuals in hospital with the disease or the proportion of the population with a marker of prior infection.
The same issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence surveys are one source of prevalence-type data that can be unbiased.
For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

However, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

@@

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Specific studies, \eg longitudinal studies that follow  individuals over time, allow  estimation of many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately powered.
Large cohort studies are expensive to run.

@@

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as resulting from a point process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
A common model is the Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
These include components of the transmission process  but do not explicitly describe the process.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions to incorporate heterogeneity, \eg through stratification.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Observational models} \label{intro:sec:disease-model}

The infection process is hardly ever directly observed.
However, as seen in previous sections, indirect observations are available on the infection's consequences.
These can be linked to the latent infection process through appropriate observational models.
At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
This equation describes the expectation of an infectious sequelae process ({\it e.g.} doctor's consultation, death) $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability distribution of the time from infection to observation.
$f$ is an example of a \emph{delay function}, and  
$f$ is commonly the pdf of a distribution, or a transformation of it.

@@

Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
Typically, this is either the time to an event or the duration of a disease stage.
Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.  This generally comes from longitudinal studies where the occurrence of the infection's sequelae is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
Concerted Action (2000). Time from HIV- 1
seroconversion to AIDS and death in 13,000 individuals before the widespread use of
highly active anti-retroviral therapy: effect of age at seroconversion and exposure
category. Lancet 355, 1131-1137.)

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, the number of tests carried out varied over time and space, and for a variety of reasons.
Reasons included: policy changes on who was eligible for testing; availability of tests; and willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests did not necessarily reflect changes in disease incidence.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.

@@

Seroprevalence surveys were conducted throughout the epidemic, the largest of which was testing donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and attempt to observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the  Coronavirus (COVID-19) Infection Survey (CIS), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

Many ad hoc studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
One of the study, which  I will use in what follows, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
In ATACCC, daily testing of a small number of individuals with short follow-up was performed.
These individuals were recruited from contacts of known infections reported to NHS Test and Trace.
The density of this sampling allows for reliable estimates of how the viral load (the amount of virus in an individual) varies over time and other quantities of interest, most importantly the duration of detectability (how long an individual is RT-PCR positive for).
However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of detectability.
However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

@@

The massive impact of the SARS-CoV-2 pandemic led to data collection on a scale never previously seen for an infectious disease.
Analysis of this data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
Therefore, a deep understanding of the potential of each data source was not possible.
In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.

In the following subsections, I introduce the most relevant related work, carried out both prior to the pandemic and during its course.

@@

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a Poisson process as the infection model and a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis; and a trivial observation model where all infections are (eventually) detected.

In (see \cref{E-inc-prev}) I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$. In addition, unlike in the original formulations, not all infections are detected by the CIS as in the CIS a sample of the population receive RT-PCR tests, not the entire population.
Therefore, I modify the observation model to include both the dependency and the sampling error arising from this process.
% The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.

@@

In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
Previous work has attempted to estimate $f$.
An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported observed duration was 83 days, although no individuals were infectious beyond day nine.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of detectability.
However, the study's cohort consisted only of healthcare workers, \ie working-age adults.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, which was successfully pursuing a zero-COVID strategy at the time, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.

\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design does not provide information on how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of detectability, because the infection's true start time would be prior to the first positive test.

For the purposes of this thesis, I require estimates of duration applicable to the general population.
As no such estimates of the full distribution of the duration are available in the literature, a key component of this work is devoted to estimating them.
To do so, I will combine the information from the ATACCC and CIS studies.
